---
title: "R for Data Science"
author: "Rutendo F. Sigauke"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r include = FALSE}
#Set options for the entire document in the first chunk.
knitr::opts_chunk$set(message = FALSE, echo=FALSE, eval = FALSE)
```

# Overview

In this worksheet we will go over essential R tools for Data Science. This document will go over the following topics:

1. We will revisit **R data structures**, and how they can be used for Data Science. 
2. We will explore useful **R commands**.
3. We will write **R functions** and execute them.
4. We will review **R loops**.
5. We will talk about communicating our data with an **R notebook** (Note: This document was generated in an R Notebook.).

# Resources

- For more details of using R for Data Science, see: https://r4ds.hadley.nz/
- For more on general R usage: https://intro2r.com/ and http://venus.ifca.unican.es/Rintro/index.html# 
- For advanced R data structures: http://adv-r.had.co.nz/Data-structures.html
- Deeper dive into R markdown: https://bookdown.org/yihui/rmarkdown/
- R markdown and R notebooks cheat sheets: https://rstudio.github.io/cheatsheets/html/rmarkdown.html
- Tidyverse style guide: https://style.tidyverse.org/index.html


# R data structures

There are various data structures in R. This includes *vectors*, *factors*, *matrices*, *arrays*, *data frames*, *lists*, and *functions*. 

- *Vectors*: are one-dimensional arrays used to store collection data of the same mode. 
- *Factors*: are vectors of categorical variables designed to group the components of another vector with the same size.
- *Matrices*: are two-dimensional arrays that can store collections of data of the same mode. Values in a matrix are accessed by two integer indices `mat[row,column]`. 
- *Arrays*: are similar to matrices but they can be multi-dimensional.
- *Lists*: are ordered collection of objects and the elements can be of different types.
- *Data Frames*: are a generalization of matrices where different columns can store different mode data (e.g. characters, numeric).
- *Functions*: are objects created by the to automate specific operations.

More details on each of the data structures can be found here: http://venus.ifca.unican.es/Rintro/dataStruct.html 

```{r}
################################
#Installing packages           #
################################
#install.packages("FactoMineR")
#install.packages("factoextra")
#install.packages("ggplot2")
#install.packages("plotly")
#install.packages("ggfortify")

```


## Vectors

1. Generate a vector of several numeric values.
2. Check the type of data class.
3. Perform a mathematical operation on the vector.
4. Create a sequence of integers.
5. Repeat value the value 42 eight times.
6. Create a random sequence of vectors (**CHALLENGE:** explain the `sample` function)

```{r}
#Generate a vector with several numeric values 
a_vector <- c(2, 10, 27, 100, 36)
a_vector

#checking data class
class(a_vector)

#perform mathematical operations on the vector 
a_vector+2

#sequence of vectors
a_seq_vectors <- c(1:8)
a_seq_vectors

#repeat the value 42 eigth times
a_rep_vectors <- rep(42, times=8)
a_rep_vectors

#random sequence of values
set.seed(42) #create a seed to ensure reproducibility
a_sample_vectors <- sample(1:8, 10, replace=TRUE) #CHALLENGE: explain the function
a_sample_vectors

#?sample

```

## Factors

1. Create a vector of factors.

```{r}

#factor vector
#Generate b vector with several characters 
b_vector <- c("A", "B", "A", "C", "A")
b_vector

b_factor_vector <- factor(b_vector)
b_factor_vector
```

## Matrices

1. Create a 3 rows and 4 columns matrix with values 1 to 12.
2. Create a second matrix with values of your choice.
3. Apply mathematics operations **between** the two matrices. See more examples here: https://www.geeksforgeeks.org/operations-on-matrices-in-r/

```{r}
#matrix with values 1 to 12. Matrix has 3 rows and 4 columns.
a_matrix <- matrix(1:12, nrow=3, ncol=4) 
a_matrix

#second matrix with 12 separate values
b_matrix <- matrix(13:24, nrow=3, ncol=4) 
b_matrix

##multiplying the two matrices (+,-,*,/)
a_matrix*b_matrix

```

## Arrays

1. Create an array with 3 dimensions.

```{r}
an_array <- array(1:24, dim=c(2,3,4))
an_array
```

## Data frames

1. Create a data frame with three columns with `id` (character), `score` (numeric), and `grade` (character).
2. Check the structure of the data frame.
3. Change the grade column to factor and view the factor.
4. Add column with rank of grades as a factor (e.g. "first","second"...) and view the factors. (**CHALLENGE:** reorder the factor so that the order makes sense (i.e. increasing))
5. Check the class of each column in the data frame


```{r}
# Create an example data frame
example_df <- data.frame(id = c("id1", "id2", "id3", "id4", "id5"),
                         score = c(90, 95, 10, 70, 10),
                         grade = c("A", "A", "F", "C", "F"))

#checking structure of df
str(example_df)

#converting grade to factor
example_df$grade <- as.factor(example_df$grade)

#viewing the factor
example_df$grade

##adding rank column
example_df$rank <- factor(c("first","first","fifth","third","fifth"))

#viewing the factor
example_df$rank ##CHALLENGE: reorder the factor so that the order makes sense (i.e. increasing)

##NOTE: changing the order of factors can come in handy when plotting and you want to change the order of variables.
example_df$rank <- factor(example_df$rank, levels = c("first","second","third","forth","fifth"))
example_df$rank

#check the column classes
sapply(example_df, class) ##what does sapply do? Alternative is to used str() as we did above.
#?sapply

```


## Lists

1. Create a list with three characters.

```{r}

a_list <- list("apple","banana","guava")
a_list

```
2. Create a list with a data.frame, list, and matrix. You can select from the data structures created above.
3. View contents of the new list one item at a time.
4. Check names of the entries in the list.
5. Rename each of the names in the list.

```{r}
#this is a list containing a data.frame, list and matrix
list_of_datastructures <- list(example_df, a_list, a_matrix)
list_of_datastructures

#viewing first item in list
list_of_datastructures[[1]]
list_of_datastructures[[2]]
list_of_datastructures[[3]]

##first check what the names are
names(list_of_datastructures)

#add new names
names(list_of_datastructures) <- c("example_dataframe","example_list","example_matrix")

#check names again
names(list_of_datastructures) #this only returns the names
list_of_datastructures #this will return previews of each list item and the names will be listed.
```

# Functions

## Simple function

1. Write a simple function that takes in a name and prints "Hello World, my name is **my_name** !"

```{r}
##function with hello world
hello_world <- function(my_name){
  paste("Hello World, my name is", my_name, "!") 
}

##running function
hello_world("Rutendo")
```

## Loops in R

Loops are helpful for performing repetative tasks. There are three types of loops `for`, `while` and `repeat`. Loops are essential for programming across many languages. 

See https://intro2r.com/loops.html and https://www.geeksforgeeks.org/loops-in-r-for-while-repeat/ for more details and examples.

### `for` loops

i. Write code to display numbers from 1 to 10 using `for` loop in R. 

```{r}
#display 1 to 10 in loop
for (i in 1:10){
  print(i)
}
```

ii. Write a `for` loop to iterate across a list. (**CHALLENGE:** explain the `seq_along` function)

```{r}

#create a list
my_list <- list("apples","bananas","pineapples","kiwi")

#loop through my list (Method 1)
for (i in my_list){
  print(i)
}

#loop through my list (Method 2)
for (i in 1:length(my_list)){
  print(my_list[[i]])
}

#loop through my list (Method 3)
for (i in seq_along(my_list)){ ##what does seq_along do?
  print(my_list[[i]])
}

```

iii. Loop across a 3x3 matrix and print each cell value.

```{r}
# Create a 3x3 matrix of integers
my_matrix <- matrix(1:9, nrow = 3, ncol=3)
 
# Loop through matrix and print each cell value
for (i in seq_len(nrow(my_matrix))) {
  for (j in seq_len(ncol(my_matrix))) {
    current_cell <- my_matrix[i, j]
    print(paste("Value in cell :", current_cell))

  }
}

# Loop through matrix and print each cell value
#and also print the row and colum of the cell
for (i in seq_len(nrow(my_matrix))) {
  for (j in seq_len(ncol(my_matrix))) {
    current_cell <- my_matrix[i, j]
    print(paste("Row",i,"and Column",j))
    print(paste("Value in cell :", current_cell))
    print("---------------------------")
  }
}
```

### `while` loop

i. Write code to display numbers from 1 to 10 using a `while` loop. 

```{r}

#initialize a variable for the value to display
val = 1

# using while loop to print 1-10
while (val <= 10){
  print(val)
  val = val + 1
}

```
### `repeat` loop

i. Write code to display numbers from 1 to 10 using a `repeat` loop. 

```{r}

#initialize a variable for the value to display
val = 1
 
#using repeat loop to print 1-10
repeat {
  #print the values
  print(val)
  val = val + 1
 
  #checking stop condition
  if(val > 10){
    #using break statement
    #to terminate the loop
    break
  }
}

```


### Alternatives to loops

There are other options other than loops that fall in the `apply` family (`apply()`, `lapply()`, `tapply()`, `sapply()`, `vapply()`, and `mapply()`). See this chapter in *An Introduction to R* https://intro2r.com/loops.html#if-not-loops-then-what for more details.

## Complex functions with loops

2. Create a function `hello_friend` that says "Hello **example_name**!".


```{r}

#function saying "Hello" to a friend.
hello_friend <- function(friend){
  paste("Hello ", friend,"!") 
}

#Creating a vector to a list of friends
my_friends <- c("Mary","Robin","Esteban","Meg",
                "Sam","Lynn","Zach","Chris")

#Running function on the whole vector
hello_friend_out <- hello_friend(my_friends)
print(hello_friend_out)


```
3. Create a vector with 4 names and loop through to run the function `hello_friend`.

```{r}
#looping through names and running the 'hello_friend' function
for (a_friend in my_friends){

  hello_friend_out <- hello_friend(a_friend)
  
	print(hello_friend_out)
  
}
```


4. Write and run a function for the clustering algorithm (from Apr 8th). (**HINT:** We can use a `for` loop across all the k values.)


```{r}

#load libraries
#I usually load my libraries at the top of the notebook and specify why they are needed.
library(ggplot2) ## package for plotting that is used in factoextra
library(factoextra) ## visualize cluster analyses


#load the data
data(USArrests)

#remove NAs from dataframe
USArrests_df <- na.omit(USArrests)

#scale the data
USArrests_df <- scale(USArrests_df)

#function that clusters and plots the different ks
f1 <- function(inputdf, c, numstarts){
	k <- kmeans(inputdf, centers = c, nstart = numstarts)
	
	p <- fviz_cluster(k, geom = "point", data = inputdf) + 
	     ggtitle(paste("k = ",c))

  print(p) #what do you see?
}

#set up a list of values from 2 to 5
num_centers <- seq(2,5,1)
num_start <- 25

#parse through the different centers and run the clustering algorithm
for (i in num_centers){
  
  f1(USArrests_df, i, num_start)
  
}

```


4. Clean up the code for clustering, and run the updated code!

```{r, results='hide'}
# 1: Running K-means
run_kmeans <- function(inputdf, numcenter, numstarts){
	# run kmeans() function with variable starts and centers
	k_starts <- kmeans(inputdf, centers = numcenter, nstart = numstarts)
	# return the results from the clustering
	return(k_starts)
}

# 2: Plotting the clusters
plot_kmeans_clusters <- function(kout, inputdf, numcenter){
	# plot the clusters 
	cluster_plot <- fviz_cluster(kout, geom = "point", data = inputdf) + 
 ggtitle(paste("k = ",numcenter))
	# return the plot
	print(cluster_plot)
}

# 3: Running the K-means clustering and plotting functions
#set up a list of values from 2 to 5
num_centers <- seq(2,5,1)
num_start <- 25

#parse through the different centers and run the clustering algorithm
for (center in num_centers){
  
  example_k <- run_kmeans(inputdf = USArrests_df,
                          numcenter = center,
                          numstarts = num_start)

  plot_kmeans_clusters(kout=example_k, 
                       inputdf = USArrests_df,
                       numcenter = center
                       )  
  
}

```

5. Write functions to save analysis files (table with clusters and figures as png).

**HINT 1:** To save tables we can use `write.table()`, and you can specify output formats (e.g. csv or tsv). 
**HINT 2:** The `ggplot2` package has a function to save the figures called `ggsave()`.

```{r, results='hide'}

#1: Create a function that saves the original table including the clusters
save_kcluster_table <- function(inputdf, kout, output_dir, output_file){
  
  #make sure the table is a dataframe
  inputdf <- as.data.frame(inputdf) 
  
  #add column with cluster informations
  inputdf$Cluster <- kout$cluster 
  
  #save file to specific folder
  write.table(inputdf,
              file = paste0(output_dir,"/",output_file),
              sep = "\t")

}

#2: Create function for saving plot as png
save_kcluster_figure <- function(kplot, output_dir, output_file){
  
  #save plot with ggplots' `ggsave` 
  ggsave(filename = paste0(output_dir,"/",output_file),
         plot = kplot,
         device = "png")

}

# 3: Running the K-means clustering and plotting functions
#set up a list of values from 2 to 5
num_centers <- seq(2,5,1)
num_start <- 25
outdir <- "/Users/rutendo/Desktop/MCDB_3450_6440/Worksheets/RforDataScience"

#parse through the different centers and run the clustering algorithm
for (center in num_centers){
  
  outtab <- paste0("USArrests_k",center,".tsv")
  outfig <- paste0("USArrests_figure",center,".png")
  
  example_k <- run_kmeans(inputdf = USArrests_df,
                          numcenter = center,
                          numstarts = num_start)

  example_plot <- plot_kmeans_clusters(kout = example_k, 
                                       inputdf = USArrests_df,
                                       numcenter = center) 
  
  save_kcluster_table(inputdf = USArrests_df, 
                      kout = example_k, 
                      output_dir = outdir,
                      output_file = outtab)
  
  save_kcluster_figure(kplot = example_plot, 
                       output_dir = outdir, 
                       output_file = outfig)
  
}


```


# Miscellaneous

1. The package `cowplot` allows for easy plotting along with `ggplot2`. Examples : https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html and https://wilkelab.org/cowplot/articles/index.html
2. We can create interactive plots with `plotly` and `ggfortify`. The `plotly` website https://plotly.com/r/basic-charts/ has a few examples.
3. `shiny` offers more versatile interactive applications for communicating data in R. The `shiny` website https://www.rstudio.com/products/shiny/ has a number of example apps and details on hosting an app locally through RStudio.



