---
title: "MnM_Summary_Notebook"
author: "Rutendo Sigauke"
date: "2025-07-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# M&M Counting Exercise 

This notebook summarizes the M&M counts from the Short Read Workshop. In this experiment,
we have two bowls (red and green) with M&Ms. The two bowls represent an experimental set up, say green is our control while red is our treated. 

The goal of this exercise is to visualize how we count reads in Short Read sequencing and ways to treat bias (i.e. contamination and batch effects).

## Loading libraries

We will be using these libraries below. 

```{r loadlibraries, echo=TRUE}
# First, install the required packages
#install.packages('data.table')
#install.packages('ggplot2')
#install.packages('cowplot)

# Load the required library 
library(data.table)
library(ggplot2)
library(cowplot)

```

## Loading data

The summary statistics are for each color of the M&M colors. What other candies do you observe in our two bowls?

```{r load_data, echo = FALSE}

# Load the data from m&m counts as a data table
# the data.table package can handle large datasets. 
# NOTE: to set the nThreads to a number you have
mnm_dt <- data.table::fread("~/Desktop/SR2025/day7/notebooks/2025_M_and_M.tsv", nThread = 1) 
dim(mnm_dt) #get the dimensions of data.table
head(mnm_dt) # getting the top lines
```

## Exploring the data

```{r plotting_hist, echo = FALSE}
# Histogram of each color, binwidth will depend on approx size of samples
ggplot(mnm_dt, aes(x=blue)) + 
  #geom_histogram(binwidth=10, fill="blue", color="gray90") +
  geom_histogram(bins=10, fill="blue", color="gray90") +
  labs(title = " ", 
       y = "Count",
       x = "Blue (N)") +
  theme_cowplot(20) +
  theme(plot.title = element_text(hjust = 0.5),
        title = element_text(size = 30), 
        axis.title = element_text(size = 20),
        axis.text.y = element_text(size = 18),
        axis.text.x = element_text(size = 18)) 

```

We can do this one M&M, skittle, e.t.c at a time. But, we can make this step
simpler. In order to plot all the M&Ms in the data.table, we have to rearrange the 
the table from a **wide** for to a **long** format. There are a couple ways we can 
achieve this. 

1. With `pivot_longer` from [tidyr](https://tidyr.tidyverse.org/reference/pivot_longer.html) or 
2. With `melt` from [reshape2](https://www.rdocumentation.org/packages/reshape2/versions/1.4.4/topics/melt).

Below, we can use the `melt` function. Additional resources for reshaping a `data.table` can be found [here](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html).

```{r wide_to_long,echo = FALSE}
# get long format for the table
mnm_long_dt <- melt(mnm_dt, id.vars = c("Name","red_bowl","virtual"),
                    measure.vars =c('red_hots','mike_ike', 'blue', 
                                    'brown', 'orange', 'yellow', 'red',
                                    'green',  'orange_sktl', 
                                    'green_sktl', 'red_sktl', 
                                    'yellow_sktl', 'purple_sktl'))

```


```{r rename_columns, echo = FALSE}
# If you have completed the CHALLENGE above, you will notice
#  the headings are generic. We can rename each of the columns below
colnames(mnm_long_dt) <- c("Name","red_bowl","virtual","candy_type","count")

```

```{r plot_multi_hist, echo = FALSE}
ggplot(mnm_long_dt, aes(count, fill=candy_type)) + 
  geom_histogram(bins=10) + 
  facet_wrap(~candy_type, scales = 'free_x') + 
  ggtitle("Raw Counts") +
  labs(title = "Raw Counts", 
       y = "Count",
       x = "M&M+ Counts",
       fill = "Candy") +
  theme_cowplot(12) +
  theme(plot.title = element_text(hjust = 0.5),
        title = element_text(size = 12), 
        axis.title = element_text(size = 10),
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 8)) 
```

## Comparing Green and Red Bowl Counts between Samples

### Split the data into respective bowls

Here, to compare between the green and red bowl, we will split the M&M counts between the two bowls.

```{r split_bowls, echo = FALSE}
# Here we are splitting the bowls with M&M 
mnm_redbowl_dt <- subset(mnm_dt, red_bowl == "TRUE")
mnm_greenbowl_dt <- subset(mnm_dt, red_bowl == "FALSE")

#summary of dimensions
print(paste0("Dimensions of Red Bowl (rows) = ",
             nrow(mnm_redbowl_dt)))
print(paste0("Dimensions of Green Bowl (rows) = ",
             nrow(mnm_greenbowl_dt)))

```

### Calculate the Fold Change on Raw Counts

**Note:** Since we have uneven number of samples/individuals between the red and green bowl, we will have to use the smallest sample sizes for both data.tables to calculate the fold change

```{r fold_change_green_red, echo = FALSE}
# 1: smallest samples size to use for fold change
minsample <- min(nrow(mnm_redbowl_dt), nrow(mnm_greenbowl_dt))

# 2: get value columns for just the M&Ms
mnm_redbowl_val_dt <- mnm_redbowl_dt[1:minsample,c('blue','brown','orange',
                                            'yellow','red','green')]

mnm_greenbowl_val_dt <- mnm_greenbowl_dt[1:minsample,c('blue','brown','orange',
                                              'yellow', 'red', 'green')]

# 3: calculate the fold change of the columns
mnm_redgreen_fc_dt <- mnm_redbowl_val_dt/mnm_greenbowl_val_dt
dim(mnm_redgreen_fc_dt)

```
We can plot histograms of the fold change calculations.

```{r blue_fc_hist, echo = FALSE}
# Histogram of each color, binwidth will depend on approx size of samples
# we can pick number of bins instead of bin width
blue_fc_hist <- ggplot(mnm_redgreen_fc_dt, aes(x=blue)) + 
                  geom_histogram(bins = 20, fill="blue", color="gray90") + 
                  labs(title = "Blue M&Ms", 
                       y = "Count",
                       x = "Red/Green Bowl Fold Change") +
                  theme_cowplot(20) +
                  theme(plot.title = element_text(hjust = 0.5, face="plain"),
                        title = element_text(size = 20), 
                        axis.title = element_text(size = 18),
                        axis.text.y = element_text(size = 15),
                        axis.text.x = element_text(size = 15)) 

blue_fc_hist
```
```{r blue_logfc_hist, echo = FALSE}
# Histogram of each color, binwidth will depend on approx size of samples
# we can pick number of bins instead of bin width
blue_logfc_hist <- ggplot(mnm_redgreen_fc_dt, aes(x=log(blue, 2))) + 
                  geom_histogram(bins = 20, fill="blue", color="gray90") + 
                  labs(title = "Blue M&Ms", 
                       y = "Count",
                       x = "log2 Fold Change \n Red Bowl/Green Bowl") +
                  theme_cowplot(20) +
                  theme(plot.title = element_text(hjust = 0.5, face="plain"),
                        title = element_text(size = 20), 
                        axis.title = element_text(size = 18),
                        axis.text.y = element_text(size = 15),
                        axis.text.x = element_text(size = 15)) 

blue_logfc_hist
```

```{r red_fc_hist, echo = FALSE}
# Histogram of each color, binwidth will depend on approx size of samples
# we can pick number of bins instead of bin width
red_fc_hist <- ggplot(mnm_redgreen_fc_dt, aes(x=red)) + 
                  geom_histogram(bins = 20, fill="red", color="gray90") + 
                  labs(title = "Red M&Ms", 
                       y = "Count",
                       x = "Red/Green Bowl Fold Change") +
                  theme_cowplot(20) +
                  theme(plot.title = element_text(hjust = 0.5, face="plain"),
                        title = element_text(size = 20), 
                        axis.title = element_text(size = 18),
                        axis.text.y = element_text(size = 15),
                        axis.text.x = element_text(size = 15)) 

red_fc_hist
```

```{r red_logfc_hist, echo = FALSE}
# Histogram of each color, binwidth will depend on approx size of samples
# we can pick number of bins instead of bin width (RED)
red_logfc_hist <- ggplot(mnm_redgreen_fc_dt, aes(x=log(red, 2))) + 
                  geom_histogram(bins = 20, fill="red", color="gray90") + 
                  labs(title = "Red M&Ms", 
                       y = "Count",
                       x = "log2 Fold Change \n Red Bowl/Green Bowl") +
                  theme_cowplot(20) +
                  theme(plot.title = element_text(hjust = 0.5, face="plain"),
                        title = element_text(size = 20), 
                        axis.title = element_text(size = 18),
                        axis.text.y = element_text(size = 15),
                        axis.text.x = element_text(size = 15)) 

red_logfc_hist
```


```{r combinig_plots, fig.dim = c(8, 4), echo = FALSE}

red_blue_logfc_hist  <- plot_grid(blue_logfc_hist,
                               red_logfc_hist, 
                               labels = c('A', 'B'),
                               label_size = 20)
red_blue_logfc_hist
```

```{r all_fc_hist, echo = FALSE}

#get list of unique candies
candies <- unique(colnames(mnm_redgreen_fc_dt))

#create an empty list to add our plots
candy_fc_hist_list <- list()

# calculate log2 fc
mnm_redgreen_logfc_dt <- log(mnm_redgreen_fc_dt, base=2)

# loop through all the different candies and generate histogram
for (i in candies){
  print("-----------------------------------------------")
  print(paste(i, " M&M"))

  candy_logfc_hist <- ggplot(mnm_redgreen_logfc_dt, aes_string(x=i)) + 
    geom_histogram(bins = 20, fill=i, color="gray90") + 
    labs(title = paste(i, "M & M"), 
         y = "Count",
         x = "Log2 Fold Change \n Red Bowl/Green Bowl") +
    theme_cowplot(15) +
    geom_vline(xintercept = 0, linetype="dashed", 
               color = "gray40", size=0.5) +
    scale_x_continuous(limits = c(-4,4)) +
    theme(plot.title = element_text(hjust = 0.5, face="plain"),
          title = element_text(size = 15), 
          axis.title = element_text(size = 12),
          axis.text.y = element_text(size = 10),
          axis.text.x = element_text(size = 10))
  
  candy_fc_hist_list[[i]] <- candy_logfc_hist
  print("-----------------------------------------------")
  
}

#plot the histograms for all candies

candy_fc_hist_all <- cowplot::plot_grid(plotlist = candy_fc_hist_list, ncol = 3)
candy_fc_hist_all
    
```

### Normalize the Counts by Sample Size

```{r normalize_mnm_counts, echo = FALSE}

# 1: we will get row totals for each sample (individual)
mnm_redbowl_rowtotal <- rowSums(mnm_redbowl_val_dt)
mnm_greenbowl_rowtotal <- rowSums(mnm_greenbowl_val_dt)

# 2: divide each row item with the row total
mnm_redbowl_val_norm_dt <- mnm_redbowl_val_dt[,.SD/mnm_redbowl_rowtotal] # using data.table sytax
mnm_greenbowl_val_norm_dt <- mnm_greenbowl_val_dt[,.SD/mnm_greenbowl_rowtotal] 

# 3: get the fold change for normalized counts
mnm_redgreen_norm_fc_dt <- mnm_redbowl_val_norm_dt/mnm_greenbowl_val_norm_dt
colnames(mnm_redgreen_norm_fc_dt) <- paste0(colnames(mnm_redgreen_norm_fc_dt),
                                            "_norm") #update the column names to show that they are normalized
head(mnm_redgreen_norm_fc_dt)

```
**Note:** were are using `cbind` from base R. Best practices is to use an ID 
maintaining approaches like `merge` from [base R](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/merge) or  
`left_join` from [dplyr](https://dplyr.tidyverse.org/reference/mutate-joins.html).

```{r compare_foldchanges, echo = FALSE}

# we can compare the raw count fold change with the normalized fold change. 
mnm_redgreen_compare_fc_dt <- cbind(mnm_redgreen_fc_dt,
                                    mnm_redgreen_norm_fc_dt)
head(mnm_redgreen_compare_fc_dt)
```

```{r plot_blue_norm_original, echo = FALSE}
#calculate log2 fold change 
mnm_redgreen_compare_logfc_dt <- log(mnm_redgreen_compare_fc_dt, base=2)

# scatterplot of log2 FC between the original and read normalized
ggplot(mnm_redgreen_compare_logfc_dt, aes(x=blue, y=blue_norm)) + 
  geom_point(shape=21, fill="blue", color="gray90", size=4) +
  labs(title = "Blue M&M \n Original Counts vs. Read Normalized Counts", 
       y = "log2 FC \n Red/Green Red Normalized",
       x = "log2 FC \n Red/Green Original") +
  geom_abline(intercept = 0, slope=1, 
              linetype='dashed', color = 'gray20', linewidth=0.5) +
  theme_cowplot(18) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 18), 
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x = element_text(size = 14)) 
```

```{r compare_all_logfc, echo = FALSE}

#get list of unique candies
candies <- unique(colnames(mnm_redgreen_fc_dt))

#create an empty list to add our plots
mnm_redgreen_compare_logfc_list <- list()

# calculate log2 fc
mnm_redgreen_logfc_dt <- log(mnm_redgreen_fc_dt, base=2)

# loop through all the different candies and generate histogram
for (i in candies){
  print("-----------------------------------------------")
  print(paste(i, " M&M"))
  
  mnm_redgreen_compare_logfc_plot <- ggplot(mnm_redgreen_compare_logfc_dt, 
                                            aes_string(x=i, y=paste0(i,"_norm"))) + 
  geom_point(shape=21, fill=i, color="gray90", size=4) +
  labs(title = paste(i, " M&M"), 
       y = "log2 FC \n Red/Green Read Normalized",
       x = "log2 FC \n Red/Green Original") +
  geom_abline(intercept = 0, slope=1, 
              linetype='dashed', color = 'gray20', linewidth=0.5) +
  theme_cowplot(8) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 12), 
        axis.title = element_text(size = 10),
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 8)) 
  
  mnm_redgreen_compare_logfc_list[[i]] <- mnm_redgreen_compare_logfc_plot
  print("-----------------------------------------------")
  
}

#plot the histograms for all candies
mnm_redgreen_compare_logfc_plot_all <- cowplot::plot_grid(plotlist = mnm_redgreen_compare_logfc_list, 
                                                          ncol = 3)
mnm_redgreen_compare_logfc_plot_all

```


### Normalize the Counts by Spike-ins 

*Note:* The number of spike-in (Red Hots, Mike and Ike) between the two bowls was kept constant.

```{r normalize_mnm_counts_spike, echo = FALSE}

# 1: we will get marshmallow totals for each sample (individual)
mnm_redbowl_spiketotal <- rowSums(mnm_redbowl_dt[,c("red_hots","mike_ike")])
mnm_greenbowl_spiketotal <- rowSums(mnm_greenbowl_dt[,c("red_hots","mike_ike")])

# 2: divide each row item with the marshmallow total
mnm_redbowl_spike_norm_dt <- mnm_redbowl_val_dt[,.SD/mnm_redbowl_spiketotal] # using data.table syntax
mnm_greenbowl_spike_norm_dt <- mnm_greenbowl_val_dt[,.SD/mnm_greenbowl_spiketotal] 

# 3: get the fold change for normalized counts
mnm_redgreen_spike_norm_fc_dt <- mnm_redbowl_spike_norm_dt/mnm_greenbowl_spike_norm_dt
colnames(mnm_redgreen_spike_norm_fc_dt) <- paste0(colnames(mnm_redgreen_spike_norm_fc_dt),
                                            "_spike") #update the column names to show that they are normalized
head(mnm_redgreen_spike_norm_fc_dt)

```

```{r compare_foldchanges_marsh, echo = FALSE}

# we can compare the raw count fold change with the normalized fold change. 
mnm_redgreen_compare_all_fc_dt <- cbind(mnm_redgreen_compare_fc_dt, 
                                        mnm_redgreen_spike_norm_fc_dt)
head(mnm_redgreen_compare_all_fc_dt)
```
```{r plot_blue_marshnorm_original, echo = FALSE}

ggplot(mnm_redgreen_compare_all_fc_dt, aes(x=blue, y=blue_spike)) + 
  geom_point(shape=21, fill="blue", color="gray90", size=4) +
  labs(title = "Blue M&M \n Original vs. Spike-in Normalized Counts", 
       y = "Red/Green Spike-in Normalized",
       x = "Red/Green Original") +
  geom_abline(intercept = 0, slope=1, 
              linetype='dashed', color = 'gray20', linewidth=1) +
  theme_cowplot(20) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 20), 
        axis.title = element_text(size = 18),
        axis.text.y = element_text(size = 16),
        axis.text.x = element_text(size = 16)) 
```


```{r compare_all_logfc_spike, echo = FALSE}
#create an empty list to add our plots
mnm_redgreen_comparespike_logfc_list <- list()

# calculate log2 fc
mnm_redgreen_compare_all_logfc_dt <- log(mnm_redgreen_compare_all_fc_dt, base=2)

# loop through all the different candies and generate histogram
for (i in candies){
  print("-----------------------------------------------")
  print(paste(i, " M&M"))
  
  mnm_redgreen_comparespike_logfc_plot <- ggplot(mnm_redgreen_compare_all_logfc_dt, 
                                            aes_string(x=i, y=paste0(i,"_spike"))) + 
  geom_point(shape=21, fill=i, color="gray90", size=4) +
  labs(title = paste(i, " M&M"), 
       y = "log2 FC \n Red/Green Spike-in Normalized",
       x = "log2 FC \n Red/Green Original") +
  geom_abline(intercept = 0, slope=1, 
              linetype='dashed', color = 'gray20', linewidth=0.5) +
  theme_cowplot(8) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 12), 
        axis.title = element_text(size = 10),
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 8)) 
  
  mnm_redgreen_comparespike_logfc_list[[i]] <- mnm_redgreen_comparespike_logfc_plot
  print("-----------------------------------------------")
  
}

#plot the histograms for all candies
mnm_redgreen_comparespike_logfc_all <- cowplot::plot_grid(plotlist = mnm_redgreen_comparespike_logfc_list, 
                                                          ncol = 3)
mnm_redgreen_comparespike_logfc_all

```
## Comparing *TOTAL* normalized vs original counts across M&Ms

```{r total_fc_original, echo = FALSE}

# 1: get total values for green to read bowl
#i- raw counts sums
mnm_redbowl_sum <- sum(mnm_redbowl_val_dt)
mnm_greenbowl_sum <- sum(mnm_greenbowl_val_dt)

#ii- spike-in counts
mnm_redbowl_spiketotal_sum <- sum(mnm_redbowl_dt[,c("red_hots","mike_ike")])
mnm_greenbowl_spiketotal_sum <- sum(mnm_greenbowl_dt[,c("red_hots","mike_ike")])

# 2: Get totals for each M&Ms 
mnm_redbowl_val_total <- t(as.data.frame(colSums(mnm_redbowl_val_dt)))
mnm_greenbowl_val_total <- t(as.data.frame(colSums(mnm_greenbowl_val_dt)))

# 3: fold change of sum between bowls
#i: Calculate the raw fold change
mnm_redgreen_sum_fc_df <- mnm_redbowl_val_total/mnm_greenbowl_val_total

#ii: Calculate the total normalized fold change
mnm_redbowl_val_sumnorm <- (mnm_redbowl_val_total/mnm_redbowl_sum) # color m&m/all m&m in red bowl
mnm_greenbowl_val_sumnorm <- (mnm_greenbowl_val_total/mnm_greenbowl_sum) # color m&m/all m&m in green bowl
mnm_redgreen_norm_fc_df <- mnm_redbowl_val_sumnorm/mnm_greenbowl_val_sumnorm 

#iii: Calculate the spike-in normalized fold change
mnm_redbowl_val_spikenorm <- (mnm_redbowl_val_total/mnm_redbowl_spiketotal_sum) # color m&m/red bowl spike
mnm_greenbowl_val_spikenorm <- (mnm_greenbowl_val_total/mnm_greenbowl_spiketotal_sum) # color m&m/green bowl spike
mnm_redgreen_spikenorm_fc_df <- mnm_redbowl_val_spikenorm/mnm_greenbowl_val_spikenorm 

# 4: merge all the data frames
mnm_all_foldchanges <- as.data.frame(rbind(mnm_redgreen_sum_fc_df,
                             mnm_redgreen_norm_fc_df,
                             mnm_redgreen_spikenorm_fc_df))
rownames(mnm_all_foldchanges) <- c("not_norm","count_norm","spike_norm")
mnm_all_foldchanges$norm <- c("Not Normalized","Count Normalized", "Spike-in Normalized")
mnm_all_foldchanges$norm <- factor(mnm_all_foldchanges$norm,
                                   levels = mnm_all_foldchanges$norm)
mnm_all_foldchanges
```

```{r total_fc_original_plot, echo = FALSE}

mnm_all_foldchanges_dt <- data.table::data.table(mnm_all_foldchanges)

#  fill="#3182bd"
# get long format for the table
mnm_all_foldchanges_long <- melt(mnm_all_foldchanges_dt,
                                 id.vars = c("norm"),
                                 measure.vars = c('blue', 'brown', 'orange',
                                                  'yellow', 'red', 'green'))
mnm_all_foldchanges_long$log_fc <- log(mnm_all_foldchanges_long$value, base=2)

ggplot(mnm_all_foldchanges_long, aes(x=norm, y=log_fc, fill=variable)) + 
  geom_point(shape=21, color="gray90", size=4) +
  labs(title = "Comparing Fold Change", 
       y = "log2 Fold Change \n Red Bowl/Green Bowl",
       x = "Normalization",
       fill = "M&M") +
  scale_fill_manual(values = unique(as.character(mnm_all_foldchanges_long$variable))) +
  theme_cowplot(16) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 16), 
        axis.title = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12))

```
## Exploring Control Fold Change

Since we do not have an easy way to get total fold change between control (i.e. *Green Bowl* vs *Green Bowl*). We will explore the and rough sampling between the two bowls such that both *Control A* and *Control B* with contain samples from both the *Green* and *Red* bowls.

```{r control_fc, echo=FALSE}

###################################
#consider randomizing the rows again
#mnm_row_random_dt <- mnm_dt[sample(nrow(mnm_dt)),]

# Split the table top vs. bottom
mnm_cntrl_dtA <- head(mnm_dt, minsample) #head(mnm_row_random_dt, minsample)
mnm_cntrl_dtB <- tail(mnm_dt, minsample) #tail(mnm_row_random_dt, minsample)

#check the split 
print("Control vs. Control")
print(table(mnm_cntrl_dtA$red_bowl))
print(table(mnm_cntrl_dtB$red_bowl))

#get M&M counts
cntrl_redbowl_val_dt <- mnm_cntrl_dtA[,c('blue','brown','orange',
                                            'yellow','red','green')]
cntrl_greenbowl_val_dt <-mnm_cntrl_dtB[,c('blue','brown','orange',
                                            'yellow','red','green')]
  
############################
# 1: get total values for green to read bowl
#i- raw counts sums
cntrl_redbowl_sum <- sum(cntrl_redbowl_val_dt)
cntrl_greenbowl_sum <- sum(cntrl_greenbowl_val_dt)

#ii- spike-in counts
cntrl_redbowl_spiketotal_sum <- sum(mnm_cntrl_dtA[,c("red_hots","mike_ike")])
cntrl_greenbowl_spiketotal_sum <- sum(mnm_cntrl_dtB[,c("red_hots","mike_ike")])

# 2: Get totals for each M&Ms 
cntrl_redbowl_val_total <- t(as.data.frame(colSums(cntrl_redbowl_val_dt)))
cntrl_greenbowl_val_total <- t(as.data.frame(colSums(cntrl_greenbowl_val_dt)))

# 3: fold change of sum between bowls
#i: Calculate the raw fold change
cntrl_redgreen_sum_fc_df <- cntrl_redbowl_val_total/cntrl_greenbowl_val_total

#ii: Calculate the total normalized fold change
cntrl_redbowl_val_sumnorm <- (cntrl_redbowl_val_total/cntrl_redbowl_sum) # color m&m/all m&m in red bowl
cntrl_greenbowl_val_sumnorm <- (cntrl_greenbowl_val_total/cntrl_greenbowl_sum) # color m&m/all m&m in green bowl
cntrl_redgreen_norm_fc_df <- cntrl_redbowl_val_sumnorm/cntrl_greenbowl_val_sumnorm 

#iii: Calculate the spike-in normalized fold change
cntrl_redbowl_val_spikenorm <- (cntrl_redbowl_val_total/cntrl_redbowl_spiketotal_sum) # color m&m/red bowl spike
cntrl_greenbowl_val_spikenorm <- (cntrl_greenbowl_val_total/cntrl_greenbowl_spiketotal_sum) # color m&m/green bowl spike
cntrl_redgreen_spikenorm_fc_df <- cntrl_redbowl_val_spikenorm/cntrl_greenbowl_val_spikenorm 

# 4: merge all the data frames
cntrl_all_foldchanges <- as.data.frame(rbind(cntrl_redgreen_sum_fc_df,
                             cntrl_redgreen_norm_fc_df,
                             cntrl_redgreen_spikenorm_fc_df))
rownames(cntrl_all_foldchanges) <- c("not_norm","count_norm","spike_norm")
cntrl_all_foldchanges$norm <- c("Not Normalized","Count Normalized", "Spike-in Normalized")
cntrl_all_foldchanges$norm <- factor(cntrl_all_foldchanges$norm,
                                   levels = cntrl_all_foldchanges$norm)
cntrl_all_foldchanges

```
```{r total_fc_original_plot_control, echo = FALSE}

cntrl_all_foldchanges_dt <- data.table::data.table(cntrl_all_foldchanges)

#  fill="#3182bd"
# get long format for the table
cntrl_all_foldchanges_long <- melt(cntrl_all_foldchanges_dt,
                                 id.vars = c("norm"),
                                 measure.vars = c('blue', 'brown', 'orange',
                                                  'yellow', 'red', 'green'))
cntrl_all_foldchanges_long$log_fc <- log(cntrl_all_foldchanges_long$value, base=2)

ggplot(mnm_all_foldchanges_long, aes(x=norm, y=log_fc, fill=variable)) + 
  geom_point(shape=21, color="gray90", size=4) +
  geom_jitter(data=cntrl_all_foldchanges_long,
             fill="gray90",
             width = 0.15,
             shape=21, color="gray20", size=4, 
             aes(x=norm, y=log_fc)) +
  annotate("text", x = 3, y = 0.5, 
           label = "Control vs. Control", 
           size = 4,
           color="gray60") +
  labs(title = "Comparing Fold Change", 
       y = "log2 Fold Change \n Red Bowl/Green Bowl",
       x = "Normalization",
       fill = "M&M") +
  scale_fill_manual(values = unique(as.character(mnm_all_foldchanges_long$variable))) +
  theme_cowplot(16) +
  theme(plot.title = element_text(hjust = 0.5, face='plain'),
        title = element_text(size = 16), 
        axis.title = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12))

```